Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-04-13 analysis/paper/paper.Rmd
writing in progress - reconciling changes

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009].\
In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].

@@

To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package. We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia. The adequacy of a road network projection methods depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Our interest is projecting impacts of forest change on wildlife, so we focused on common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models). Our R package provides a comparable alternative to Hardy QGIS and Landis plugins, more easily integrated into SpaDES or other R workflows. 
- as expected, none of these algorithms are great at projecting exact road locations. we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield shows better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. In particular, none of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 
- Aspatial overall performance metrics can be difficult to interpret. To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors. Better harvest & existing road data would allow us to distinguish errors in projections from errors in the input data (cite lit in coops paper)
 - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing. Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm. Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this. But note that these algorithms can get the right density without replicating the pattern or mechanism. Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. Projected wiggles are quite different from observed switchbacks. 
- If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do. But before investing in more complexity recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates (). In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). If our wildlife response models depend on distance to road or buffered disturbance footprint, these somewhat crude road projections are likely good enough.
- A related point is that it may helpful to recognize that some road metrics are easier to project than others when constructing model to project wildlife responses to disturbance ... (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.

However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. 
-->
```
<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->

>>>>

The impacts of roads on wildlife are complex and varied. Mortality risk and stress on or near roads can be elevated due to vehicle collisions, hunting pressure, predation, noise and other stressors, but roads can also provide food, facilitate movement, and provide refuge [@hill_review_2021; @trombulak_review_2000; @fahrig_effects_2009]. In Canadian forests, there is ample evidence of linear features attracting predators such as bears (*Ursus americanus*, *Ursus arctos*) and wolves (*Canis lupus*) seeking food or easy movement, while repelling prey such as woodland caribou (*Rangifer tarandus caribou*) and moose (*Alces americanus*) [@dickie_corridors_2020; @polfus_identifying_2011; @mumma_predation_2018; @beauchesne_disentangling_2013; @leblond_avoidance_2013; @laurian_interactions_2012]; there is also evidence of predators avoiding features heavily used by humans [@muhly_human_2011; @whittington_towns_2022; @proctor_effects_2020]. Changes in mortality risks, habitat quality, and connectivity associated with anthropogenic disturbance can in turn alter use and movement patterns [@tucker_moving_2018; @apps_factors_2006; @proctor_effects_2020], demography and population viability [@johnson_science_2020; @decesare_linking_2014; @barrientos_lost_2021; @apps_spatial_2013; @pinard_calving_2012; @fahrig_effects_2009; @lochhead_linking_2022; @apps_spatial_2013; @proctor_effects_2020], and genetics [@habrich_varying_2021]. Roads can also facilitate spread of invasive species [@riitters_landscape_2017], and enable other development activities [@johnson_growth-inducing_2020].

@@

To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package. We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia. The adequacy of a road network projection methods depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose. Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].

@@

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools designed for this purpose have not been available. Hardy's

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models). Our R package provides a comparable alternative to Hardy QGIS and Landis plugins, more easily integrated into SpaDES or other R workflows. 
- as expected, none of these algorithms are great at projecting exact road locations. we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield shows better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. In particular, none of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 
- Aspatial overall performance metrics can be difficult to interpret. To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors. Better harvest & existing road data would allow us to distinguish errors in projections from errors in the input data (cite lit in coops paper)
 - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing. Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm. Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this. But note that these algorithms can get the right density without replicating the pattern or mechanism. Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. Projected wiggles are quite different from observed switchbacks. 
- Focussed here on forest resource roads. Note the algorithms are also likely to work for other situations where the objectives of road building are to access target locations and minimize cost. Not a useful algorithm for projecting linear features that were constructed for some other purpose such as seismic testing (e.g. Figure \@ref(fig:ftNelson)), or for circumstances where information about target locations is not available. When considering transferring these results to other contexts, also note that in this example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. It would be interesting to test performance in a circumstance where target locations are far from existing roads. 
- If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do. But before investing in more complexity recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates (). In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). For wildlife response models informed by distance to road or buffered disturbance footprint, somewhat crude road projections are likely good enough. When constructing new wildlife response models intended for projection, it be helpful to select predictor variables such as distance to road, road presence within grid cells, and disturbance footprint that are more easily projectable than road density (Bodner).
- There are a bewildering array of possible metrics of connectivity and fragmentation we did not consider in this paper. In general, performance will vary among metrics, and it is difficult to tell what is good enough without reference to particular circumstances (objectives, species, landscapes). Rather than expanding the array of road performance metrics considered, we believe our efforts are better spent making the road projection tools easy to use, to enable users devise performance tests appropriate to their particular circumstances and goals. The best performance metrics will be those most closely linked to the outcome of interest. For example, where the outcome of interest is space use or connectivity projected from more (e.g. Whittington, Labadie) or less (samc, pither etc) species-specific movement models consider testing sensitivity of these models to the difference between observed and retrospectively projected roads. We hope R package and open workflow associated with this paper can help enable retrospective projection and relevant performance assessment in other contexts. However, we also recognize there is work associated with acquiring data to construct cost surfaces & retrospectively project roads. For those seeking a lower effort option, we also provide our output maps and an example script () to enable calculation of alternative performance metrics of interest for our Revelstoke example.

```
<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->
